Continuous Deductive Reasoning over Streamed RDF Data

### 2015-11-17 article.tex
fixed definitions to include stream union and cascaded that fix, and to be based on (sliding) windows rather than pairings of streams and ranges (allows incorporation of valid windows into definitions at the end)

A {\em stream} $S$ is an unbounded sequence of fact instances. We
define a sequence as a mapping from fact instances to instants, where
the instant records the time at which the fact instance appears on the
stream. This has the consequence that a particular fact instance
(i.e. a fact accompanied by a particular instant that is its time
annotation) may appear only once in a given stream. For convenience,
we indicate that a fact instance $\langle f, t\rangle$ appears in $S$
by writing $\langle f, t \rangle \in S$. We indicate that a fact
instance $\langle f_i, t_i \rangle$ appears strictly before a fact
instance $\langle f_j, t_j \rangle$ on stream $S$ (i.e.
$S\left(\langle f_i, t_i \rangle\right) < S\left(\langle f_j, t_j \rangle\right)$)
by writing $\langle f_i, t_i \rangle \prec \langle f_j, t_j \rangle$.

@@

@@

We initially take a {\em Continuous Datalog program} (henceforth a
C-Datalog program) $\Pi^C = \langle R, F^0, S \rangle$ to be a tuple
which comprises a finite set of rules $R$, a finite set of persistent/static
fact instances $F^0$ (all of which have the instant $0$, meaning that they
cannot appear in any state of any window, by
\refdef{continuous datalog: window} and Axioms~\ref{axiom:continuous datalog: window range leq 0} and
\ref{axiom:continuous datalog: window start leq 0}) and an in-order stream
of fact instances $S$.  It should be noted that, in practice, data may be
drawn from more than one source, and so $S$ may be a union of
the mappings represented by $n$ in-order streams $S_0,\dots,S_n$ without altering
the semantics presented here.

@@

However, this definition of a C-Datalog program does not take into
account any notion of windowing over the stream $S$, and so any fact of
which an instance occurs in $S$ will be considered true from the instant
of its occurrence in $S$.  While this is technically acceptable in the
case of sets of rules without constraints, if a constraint exists and its
body becomes satisfied by some $\Pi^C$ at some instant $i$ then the
program will be unsatisfiable from that instant on, making $E^S$ empty
from then on.  As such, we define a {\em Windowed Continuous Datalog program}
$\Pi^{C, r} = \langle R, F^0, \{\langle S_0, r_0 \rangle,\dots,\langle S_n, r_n \rangle\} \rangle$
(where $r = max(r_0,\dots,r_n)$) to be a tuple with components $R$ and $F^0$ as defined above for a
C-Datalog program, but with each part of $S = S_0 \cup \dots \cup S_n$
associated with the range $r_x \in \mathbb{N}$ of a sliding window over
$S_x$, for $x \in [0,n]$. This specifies that the fact instances
from $S_x$ are to contribute to the entailments of $\Pi^{C,r}$ from the instant
they occur in $S_x$ for a duration of $r_x$ consecutive instants.

@@

We can reduce a WC-Datalog program $\Pi^{C,r} = \langle R, F^0,
\{ \langle S_0, r_0 \rangle,\dots,\langle S_n, r_n \rangle \} \rangle$
at an instant $i \in \mathbb{T}$ to its TA-Datalog equivalent by
applying the rules $R$ to the union of the persistent facts $F^0$ and all
finite states $W^{S_x,r_x}_i$ of windows $W^{S_x,r_x}$ over streams $S_x$ of range
$r_x \in \mathbb{N}$, for all $x \in [0,n]$; we refer to this as the {\em instantaneous state} of
the WC-Datalog program, written
$\Pi^{C,r}_i = \langle R, F^0 \cup W^{S_0,r_0}_i \cup \dots \cup W^{S_n,r_n}_i \rangle$.
As previously shown in \refdef{continuous datalog: ta datalog program}, this may then be further
reduced to the Datalog program
$\Pi = \langle R, strip(F^0 \cup W^{S_0,r_0}_i \cup \dots \cup W^{S_n,r_n}_i) \rangle$.

@@

The decomposition of streams into more sparse streams without loss of
information allows us to define the state of a valid window
$W^{P \setminus N}$ as a union of the states of sliding windows over
each of the singleton streams $S_{\langle f, t \rangle} =
(\langle f, t \rangle \mapsto t^\prime)$, where
$S(\langle f, t \rangle) = t^\prime$.  The range of the sliding
window over each $S_{\langle f, t \rangle}$ is then the difference between
the instant at which $\langle f, t \rangle$ occurs in $P$ and that
at which it occurs in $N$:
\[
  W^{P \setminus N}_i = \{ \langle f, t \rangle \mid \langle f, t \rangle \in W^{S_{\langle f, t \rangle},N(\langle f, t \rangle) - P(\langle f, t \rangle)}_i\}
\]
Treated this way, the entailment of one WC-Datalog program, e.g. that in
\refeqn{continuous datalog: example WC-Datalog program},
may then be taken as a part or all of the data in a successive WC-Datalog
program, without modification of the existing semantics or loss of information
regarding the validity of the entailments.  However, in the interest
of being concise, such a successive WC-Datalog program may express valid
windows alongside any sliding windows. For example, 
a program composed of the set of rules $R^\prime \neq R$, the set of base facts
$F^{0\prime} \neq F^0$, a sliding window of range $x \geqslant r$ over the
in-order stream $S^\prime \neq S$, and the entailment of
$\Pi^{C,r} \models \langle E^0, P, N \rangle$, would be expressed as in \refeqn{continuous datalog: successive WC-Datalog program}.
\begin{gather}
  \labeleqn{continuous datalog: example WC-Datalog program}
    \Pi^{C,r} \langle R, F^0, \{W^{S,r}\} \rangle \models \langle E^0, P, N \rangle \\
  \labeleqn{continuous datalog: successive WC-Datalog program}
    \Pi^{C,x} = \langle R^\prime, F^{0\prime} \cup E^0, \{ W^{P \setminus N}, W^{S^\prime,x} \} \rangle \models \langle E^{0\prime}, P^\prime, N^\prime \rangle
\end{gather}

In conclusion, given a Continuous Datalog program $\Pi^C = \langle R,
F^0, S_0 \cup \dots \cup S_n \rangle$, we can apply to it a set of
windows of range $r_0,\dots,r_n$ to yield a Windowed Continuous Datalog
program $\Pi^{C,r} = \langle R, F^0, \{ W^{S_0, r_0}, \dots,
W^{S_n, r_n} \} \rangle$. We can evaluate these windows at
an instant $i \in \mathbb{T}$ to yield a Time-Annotated Datalog program
$\Pi^T = \langle R, F^0 \cup W^{S_0,r_0}_i \cup \dots \cup W^{S_n,r_n}_i
\rangle$, from we which can then strip the time annotations to yield a
conventional Datalog program $\Pi = \langle R, F \rangle$. The
relationship between these \ast-Datalog programs and their
entailments is expressed in the commutative diagram in
\reffig{CD-TD-D}.

>>>>

A {\em stream} $S$ is an unbounded sequence of fact instances. We
define a sequence as a mapping from fact instances to instants, where
the instant records the time at which the fact instance appears on the
stream. This makes the stream a functional mapping from fact instances
to instants, with the consequence that a particular fact instance
% (i.e. a fact accompanied by a particular instant that is its time annotation)
may appear only once in a given stream. For convenience,
we indicate that a fact instance $\langle f, t\rangle$ appears in $S$
by writing $\langle f, t \rangle \in S$. We indicate that a fact
instance $\langle f_i, t_i \rangle$ appears strictly before a fact
instance $\langle f_j, t_j \rangle$ on stream $S$ (i.e.
$S\left(\langle f_i, t_i \rangle\right) < S\left(\langle f_j, t_j \rangle\right)$)
by writing $\langle f_i, t_i \rangle \prec \langle f_j, t_j \rangle$.

@@

A \emph{union} of the mappings of two streams, written $S \cup S^\prime$,
may itself be a stream, but only in the case that each fact instance occurs
in only one of the two streams, or is mapped to the same instant in both,
thereby preserving the functionality of the mapping.  It should be noted that
the union of two in-order streams is always a stream, as each fact instance
is mapped \emph{in each in-order stream} to its inherent instant.

@@

We initially take a {\em Continuous Datalog program} (henceforth a
C-Datalog program) $\Pi^C = \langle R, F^0, S \rangle$ to be a tuple
which comprises a finite set of rules $R$, a finite set of persistent/static
fact instances $F^0$ (all of which have the instant $0$, meaning that they
cannot appear in any state of any window, by
\refdef{continuous datalog: window} and Axioms~\ref{axiom:continuous datalog: window range leq 0} and
\ref{axiom:continuous datalog: window start leq 0}) and an in-order stream
of fact instances $S$.  It should be noted that, in practice, data may be
drawn from more than one source, and so $S$ may be a union of $n$ in-order
streams $S_1,\dots,S_n$ without altering the semantics presented here.

@@

However, this definition of a C-Datalog program does not take into
account any notion of windowing over the stream $S$, and so any fact of
which an instance occurs in $S$ will be considered true from the instant
of its occurrence in $S$.  While this is technically acceptable in the
case of sets of rules without constraints, if a constraint exists and its
body becomes satisfied by some $\Pi^C$ at some instant $i$ then the
program will be unsatisfiable from that instant on, making $E^S$ empty
from then on.  As such, we define a {\em Windowed Continuous Datalog program}
$\Pi^{C, r} = \langle R, F^0, \{ W^{S_1, r_1},\dots, W^{S_n, r_n} \rangle\} \rangle$
(where $r = max(r_1,\dots,r_n)$) to be a tuple with components $R$ and $F^0$ as defined above for a
C-Datalog program, but with each part of $S = S_1 \cup \dots \cup S_n$
associated with the range $r_x \in \mathbb{N}$ of a sliding window over
$S_x$, for $x \in [1,n]$. This specifies that the fact instances
from $S_x$ are to contribute to the entailments of $\Pi^{C,r}$ from the instant
they occur in $S_x$ for a duration of $r_x$ consecutive instants.

@@

We can reduce a WC-Datalog program $\Pi^{C,r} = \langle R, F^0,
\{ W^{S_1, r_1},\dots, W^{S_n, r_n} \} \rangle$
at an instant $i \in \mathbb{T}$ to its TA-Datalog equivalent by
applying the rules $R$ to the union of the persistent facts $F^0$ and all
finite states $W^{S_x,r_x}_i$ of windows $W^{S_x,r_x}$ over streams $S_x$ of range
$r_x \in \mathbb{N}$, for all $x \in [1,n]$; we refer to this as the {\em instantaneous state} of
the WC-Datalog program, written
$\Pi^{C,r}_i = \langle R, F^0 \cup \mathop{\cup}\limits_{x=1}^n W^{S_j,r_j}_i \rangle$.
As previously shown in \refdef{continuous datalog: ta datalog program}, this may then be further
reduced to the Datalog program
$\Pi = \langle R, strip(F^0 \cup \mathop{\cup}\limits_{x=1}^n W^{S_j,r_j}_i \rangle$.

@@

The decomposition of any stream into any number of more
sparse streams without loss of information allows us to define the
state of a valid window $W^{P \setminus N}$ as a union of the states
of sliding windows over each of the singleton streams
$(\langle f, t \rangle \mapsto t)$, the union of which is equal to $P$,
where $P(\langle f, t \rangle) = t$ means that $P$ maps
$\langle f, t \rangle$ to $t$.  The range of the sliding
window over each singleton stream is the difference between
the instant at which $\langle f, t \rangle$ occurs in $P$ ($t$, as $P$
is in-order) and that at which it occurs in $N$:
\[
  W^{P \setminus N}_i = \left\{ \langle f, t \rangle \,\middle\vert\, \begin{split}
    P(\langle f, t \rangle) = t \land \\
    \langle f, t \rangle \in W^{\left( \langle f, t \rangle \mapsto t \right),N(\langle f, t \rangle) - t}_i
  \end{split}\right\}
\]
Treated this way, the entailment of one WC-Datalog program, e.g. that in
\refeqn{continuous datalog: example WC-Datalog program},
may then be taken as a part or all of the data in a successive WC-Datalog
program, without modification of the existing semantics or loss of information
regarding the validity of the entailments.  However, in the interest
of being concise, such a successive WC-Datalog program may express valid
windows alongside any sliding windows. For example, 
a program composed of the set of rules $R^\prime \neq R$, the set of base facts
$F^{0\prime} \neq F^0$, a sliding window of range $x \geqslant r$ over the
in-order stream $S^\prime \neq S$, and the entailment of
$\Pi^{C,r} \models \langle E^0, P, N \rangle$, would be expressed as in \refeqn{continuous datalog: successive WC-Datalog program}.
\begin{gather}
  \labeleqn{continuous datalog: example WC-Datalog program}
    \Pi^{C,r} \langle R, F^0, \{W^{S,r}\} \rangle \models \langle E^0, P, N \rangle \\
  \labeleqn{continuous datalog: successive WC-Datalog program}
    \Pi^{C,x} = \langle R^\prime, F^{0\prime} \cup E^0, \{ W^{P \setminus N}, W^{S^\prime,x} \} \rangle \models \langle E^{0\prime}, P^\prime, N^\prime \rangle
\end{gather}

In conclusion, given a Continuous Datalog program $\Pi^C = \langle R,
F^0, S_0 \cup \dots \cup S_n \rangle$, we can apply to it a set of
windows of range $r_0,\dots,r_n$ to yield a Windowed Continuous Datalog
program $\Pi^{C,r} = \langle R, F^0, \{ W^{S_0, r_0}, \dots,
W^{S_n, r_n} \} \rangle$. We can evaluate these windows at
an instant $i \in \mathbb{T}$ to yield a Time-Annotated Datalog program
$\Pi^T = \langle R, F^0 \cup W^{S_0,r_0}_i \cup \dots \cup W^{S_n,r_n}_i
\rangle$, from we which can then strip the time annotations to yield a
conventional Datalog program $\Pi = \langle R, F \rangle$. The
relationship between these $\ast$-Datalog programs and their
entailments is expressed in the commutative diagram in
\reffig{CD-TD-D}.
